Research Experience

### 2010-03-03 doc/thesis/thesis.tex
some more work on the thesis

\section*{Current Techniques}
\subsection*{Basic T-Test}
The most basic method of analyzing FMRI data is through a basic T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
active period separately then treating them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the likelihood that the samples came from the same distribution
(the p-value). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
5000 or more samples is on average going to generate $.05*5000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection*{General Linear Model}

@@

Secondly, these methods are still merely t-tests, which notoriously
lack robustness to non-Gaussian noise. While different techniques 
exist for imposing Gaussianity, those techniques are incapable of
discriminating noise from signal, otherwise analyzing FMRI would be
relatively simple. There is no way to know how much signal is removed
by various smoothing techniques, or even if entire regions have been
smoothed into oblivion. Instead of extensively filtering data to 
remove noise, the analysis method itself must be robust to noise,
which is why we propose here the use of particle filters.

\section*{Proposed Approach}
\subsection{Modeling}
The short term goal of current neurological models is to
learn something about the system being modeled.
While control (for instance in prosthetics), or predication
(such as BlueBrain) may some day be feasible, in the
short run we still have a great deal to learn about how our
brain works. By learning about specific systems (the dynamics
of a single person's brain), we may be able to diagnose diseases,
find their causes and even someday rectify the cause. Of course
by learning about various specific systems we can also learn 
about the general way our brains work. The reason then, for 
studying the BOLD model then is
to be able to derive the various system parameters and understand
why one brain region (say a voxel) reacts in one way, when another
acts completely differently.

>>>>

\section*{Current Techniques}
\subsection*{Basic Statistical Parametric Mapping}
The most basic method of analyzing FMRI data is through a basic T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
active period separately then treating them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the likelihood that the samples came from the same distribution
(the p-value). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
5000 or more samples is on average going to generate $.05*5000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection*{General Linear Model}
The most used form of FMRI analysis is still based on Statistical Parametric
Mapping, but is able to account for several different levels or types
of activation. By adding a General Linear Model to the analysis, the
output signal timeseries (what FMRI detects) is regressed over the weighted
sum of various confound timeseries. The equation for a general linear
model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of stimuli, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then is merely a gradient descent search to minimize the
mean squared error.

<Image of GLM>

Of course, a square wave input is not going to result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth $X(t)$ through time, and bandlimit the input. The best technique
is convolving the stimulus input with a hemodynamic response function,
which mimicks the basic shape of BOLD activation, including a delay
due to rise time and fall time. This hemodynamic signal is static however,
so every region of the brain still gets matched to the same signal,
although the weights of various stimulus or confounds (such as attention)
are allowed to vary.

<Image of Hemodynamic Response Function>

Ultimately, activation due to a particular stimuli is decided by the 
$\beta$ value corresponding to that stimuli's column of $X(t)$.
<Need to check this>
The null hypothesis as to whether the outcome was random is then
based on a t-test of the $\epsilon(t)$ timeseries.

@@

Secondly, these methods are still merely t-tests, which notoriously
lack robustness to non-Gaussian noise. While different techniques 
exist for imposing Gaussianity, those techniques are incapable of
discriminating noise from signal. There is no way to know how much 
signal is removed
by various smoothing techniques, or even if entire regions have been
smoothed into oblivion. Instead of extensively filtering data to 
remove noise, the analysis method itself must be robust arbitrary
types of noise, which is why we propose here the use of particle filters.

\section*{Proposed Approach}
\subsection{Goal}
The ultimate goal of modeling neural activation is to learn something
new about how the brain works. The General Linear Model, and 
SPM's goal is to find a correlation between input, or what a 
person is thinking, and which regions are active. This is an 
important task, but it misses the potential of FMRI. FMRI's 
greatest weakeness as a tool for studying neural activity is that
it is connected to neurons through a long chain of events. But
what if, in addition to seeing which brain regions are active
for which stimuli, we could also study pathologies such as poor
blood flow, or derive physical parameters such as blood vessel
compliance? By using a realistic physical model as the basis for
discovering the relation between the outside world and FMRI 
signal, the path to detecting activation gives all those other
things for free, precisely \emph{because} FMRI is the result of such
a long chain of events. It is actually quite common to use models to derive
the values of physiological parameters. For instance, blood volume
is often measured by injecting a dye into the blood stream and then
assuming that the dye is removed at a rate proportional to the
concentration in the blood. By performing a simple regression it is
possible to estimate blood volume quite accurately. While for
this particular project I don't claim that the results will give 
physiologically plausible results, with some tuning of the algorithm
it would very possible.

While the General Linear Model has historically done well with
detecting activation, it is limited to scaling the entire input
signal up or down. This leaves out even basic effects such as 
longer or shorter time constants, as well as nonlinear effects which
are known to be present in the BOLD system. 
**
